---
editor_options: 
  chunk_output_type: console
---

# Matching {#chapter-matching}

::: {.rmdtip}
**match**  
*verb*  
1. correspond or cause to correspond in some essential respect; make or be harmonious.  
2. be equal to (something) in quality or strength.
:::


As the name suggests, propensity score matching is concerned with matching treatment to control observations...


Matching Methods

### Nearest Neighbor Matching

MatchIt::matchit(method = "nearest")

### Optimal Pair Matching 

MatchIt::matchit(method = "optimal")

### Optimal Full Matching 

MatchIt::matchit(method = "full")

### Generalized Full Matching

MatchIt::matchit(method = "quick")

### Genetic Matching

MatchIt::matchit(method = "genetic")

X <- cbind(age, educ, black, hisp, married, nodegr, re74, re75, u74, u75)
BalanceMatrix <- cbind(age, I(age^2), educ, I(educ^2), black, hisp,
    married, nodegr, re74, I(re74^2), re75, I(re75^2), u74, u75,
    I(re74 * re75), I(age * nodegr), I(educ * re74), I(educ * re75))
gen1 <- GenMatch(Tr = Tr, X = X, BalanceMatrix = BalanceMatrix,
    pop.size = 1000)

### Exact Matching

MatchIt::matchit(method = "exact")

### Coarsened Exact Matching

MatchIt::matchit(method = "cem")

### Subclassification

MatchIt::matchit(method = "subclass")

### Cardinality and Profile Matching

MatchIt::matchit(method = "cardinality")



```{r}
lr_out <- glm(lalonde.formu, 
			  data = lalonde,
			  family = binomial(link = logit))
lalonde$lr_ps <- fitted(lr_out)  # Propensity scores
```


```{r introduction-logistic, eval=TRUE, echo=FALSE, fig.height=4, fig.cap='Propensity Scores from Logistic Regression with Sample of Matched Pairs'}
lalonde_match <- Match(Y = lalonde$re78, 
					   Tr = lalonde$treat, 
					   X = lalonde$lr_ps, 
					   M = 1,
					   caliper = 0.1,
					   replace = TRUE,
					   estimand = 'ATE')
lalonde_match_df <- data.frame(treated.ps = lalonde[lalonde_match$index.treated,]$lr_ps,
						       control.ps = lalonde[lalonde_match$index.control,]$lr_ps,
							   treated.y = 1,
							   control.y = 0)
lalonde_match_df <- lalonde_match_df[order(lalonde_match_df$control.ps),]
rows <- (1:nrow(lalonde_match_df) - 1) %% floor(nrow(lalonde_match_df) / 5) == 0
ggplot(lalonde, aes(x = lr_ps, y = treat)) + 
	geom_point(alpha = 0.5) +
	geom_smooth(method = glm, formula = y ~ x, method.args = list(family=binomial(link='logit')), se = FALSE) +
	xlim(c(0,1)) + 
	xlab('Propensity Score') + ylab('Treatment') +
	geom_segment(data = lalonde_match_df, 
				 aes(x = treated.ps, xend = control.ps, y = treated.y, yend = control.y),
				 color = 'purple', alpha = 0.1)
```
	

## One-to-One Matching ATE

One-to-one matching with replacement (the `M = 1` option). Estimating the treatment effect on the treated (default is ATT).

```{r}
rr_att <- Match(Y = lalonde$re78, 
				Tr = lalonde$treat, 
				X = lalonde$lr_ps,
				M = 1,
				estimand='ATT')
summary(rr_att) # The default estimate is ATT here
```

### Checking Balance

```{r, fig.cap='Covariate Balance Plot for Matching', fig.width=12, fig.height=8}
rr_att_mb <- psa::MatchBalance(
	df = lalonde,
	formu = lalonde.formu,
	formu.Y = update.formula(lalonde.formu, re78 ~ .),
	index.treated = rr_att$index.treated,
	index.control = rr_att$index.control,
	tolerance = 0.25,
	M = 1,
	estimand = 'ATT')
plot(rr_att_mb)
# ls(rr_att_mb)
summary(rr_att_mb)

```


```{r, eval=FALSE, echo=FALSE}
# TODO: Decide whether to include this. Could write a function to present results as a table
rr_att_mb2  <- Matching::MatchBalance(
	formul = lalonde.formu,
	data = lalonde, 
	match.out = rr_att,
	nboots = 500, 
	print.level = 0)
# rr_att_mb2 |> print() # Output is too long
```

## One-to-One matching (ATT)

```{r}
rr.ate <- Match(Y = lalonde$re78, 
				Tr = lalonde$treat, 
				X = lalonde$lr_ps,
				M = 1,
				estimand = 'ATE')
summary(rr.ate)
```

## One-to-Many Matching (ATT)


```{r}
rr2 <- Match(Y = lalonde$re78, 		
			 Tr = lalonde$treat, 
			 X = lalonde$lr_ps,
			 M = 1, 
			 ties = TRUE, 
			 replace = TRUE,
			 estimand = 'ATT')
summary(rr2) # The default estimate is ATT here
```

## The `MatchIt` Package

```{r}
matchit.out <- MatchIt::matchit(lalonde.formu, data = lalonde)
summary(matchit.out)

# Same as above but calculate average treatment effect
rr.ate <- Match(Y = lalonde$re78, 
				Tr = lalonde$treat, 
				X = lalonde$lr_ps,
				M = 1,
				ties = FALSE, 
				replace = FALSE, 
				estimand='ATE')
summary(rr.ate) # Here the estimate is ATE

## Genetic Matching
rr.gen <- GenMatch(Tr = lalonde$treat, 
				   X = lalonde$lr_ps, 
				   BalanceMatrix = lalonde[,all.vars(lalonde.formu)[-1]],
				   estimand = 'ATE', 
				   M = 1, 
				   pop.size = 16,
				   print.level = 0)
rr.gen.mout <- Match(Y = lalonde$re78, 
					 Tr = lalonde$treat, 
					 X = lalonde$lr_ps,
					 estimand = 'ATE',
					 Weight.matrix = rr.gen)
summary(rr.gen.mout)

## Partial exact matching
rr2 <- Matchby(Y = lalonde$re78, 
			   Tr = lalonde$treat, 
			   X = lalonde$lr_ps, 
			   by = factor(lalonde$nodegr),
			   print.level = 0)
summary(rr2)

## Partial exact matching on two covariates
rr3 <- Matchby(Y = lalonde$re78, 
			   Tr = lalonde$treat, 
			   X = lalonde$lr_ps, 
			   by = lalonde[,c('nodegr','married')],
			   print.level = 0)
summary(rr3)
```


